Stemmer Tool Manual

### 2018-03-14 tex/merf/OldSubmission/motivation.tex
old submission files

We motivate \framework with the example of Figure~\ref{fig:intromotiv} 
that shows how to extract direction entities from sample Arabic text. 
The text includes directions to a shopping mall and the 
figure presents a transliteration and an English translation of the text.
%The text contains some details that are not interesting to the direction extraction task. 
Interesting entities, such as names, places, relative directions, 
and numerical terms are highlighted with boxes in the text. 
We are interested in detecting those entities and presenting 
the direction as relational entities as 
shown in Figure~\ref{fig:intromotiv}.

@@

{\bf Entity detection.~}
As described in the table of Figure~\ref{fig:motiv}, the user denotes 
the ``name of person'' entities with formula $N$ 
which requires the {\em category} feature in the morphological solution of a word
to be {\tt name of person}. 
The entities $n_1,n_2,$ and $n_3$ are matches of the formula $N$. 
Similarly, the user specifies formula $P$ to denote ``name of place'' entities. 
The user specifies formula 
$R$ to denote ``relative position'' entities, and requires the stem feature to belong 
to a selected list of 
stems containing \RL{fy} and \RL{qrb}. 
Similarly, $U$ denotes numerical terms and is a disjunction of constraints 
requiring the stem feature to belong to a set of stems such as
\RL{'wl}(first), \RL{_tAny}(second), \dots \RL{`A^sr}(tenth).

\framework calls an inhouse open source morphological 
analyzer~\cite{ZaMaColing2012DemosSarf} and computes matches of 
the formulae $N$, $P$, $R$, and $U$. 
We refer to all {\em other} words in the text that do not match 
a user defined formula as {\em null} words and we denote them by $O$. 
The resulting matches are illustrated with boxes and superscripts 
in Figure~\ref{fig:intromotiv} and are listed in the table of 
Figure~\ref{fig:motiv}.

@@

The user tries several sequences of the above entities in the 
editor and checks their matches in the visualizer. 
Finally, the user is satisfied with the matches of an expression such as 
$(P|N)\!+~O?~R~O^\wedge\!2~(P|N|U)+$ where 
$|,+,?,$ and $^\wedge k$ denote disjunction, one or more, zero or one, and
up to $k$ matches, respectively.
The expression specifies a sequence of places or names of persons, 
optionally followed by a null word, 
followed by one relative position, followed by up to two possible null words, 
followed by one or more match of name of place, name of person, or numerical term. 
$O?$ and $O^\wedge 2$ are used in the expression to allow for flexible matches.
The user reaches the satisfying matches by experimenting with the visualizer and the expression editor which do not require knowledge and expertise in regular expressions.

The match trees in Figure~\ref{fig:motiv} illustrate two matches of the expression 
computed by \framework.
The first match tree refers to the text 
\RL{brj _hlyfT bAlqrb mn AltqA.t` Al-'wl}(Khalifa Tower next to the first intersection). 
The second match tree refers to the text 
\RL{dby mwl `l_A mqrbT mn h_dA Almbn_A}(Dubai Mall is located near this building).
The leaf nodes of the trees are entities and the edges and internal nodes are text, 
morphology-based, 
and word distance based relational entities. 
%The sequence and structure gives us the text of a parent node from the children nodes. 
%The same sequence can also give us the interesting (matching) morphological features. 
%The word distance is defined and abstracted by the matches of the regular expression 
%operators (internal nodes).

{\bf User defined direction relations.~}
The user now uses the relation editor to declare user defined relations that relate 
parts of the matches of the expression with each other. 
%The aim of the semantic relations to be defined is to construct the entity-relation 
%graph shown in Figure~\ref{fig:intromotiv}. 
Intuitively, 
the user wants to create relations between places, names, and numerical entities. 
A relation between two entities can be a prepositional entity.
For example, Figure~\ref{fig:intromotiv} shows the entities \RL{dby mwl}(Dubai Mall) and 
\RL{Almbn_A}(the building) related by the preposition \RL{mqrbT}(near).

Let $e_1,o_1,r,o_2,$ and $e_2$ be the 
$(P|N)+$,$O?$,$R$,$O\wedge 2$, and $(P|N|U)+$ subexpressions, respectively.
The user selected $(P|N)+$ to be an entity after noticing in the visualizer  
that it happens to capture 
non-separated sequences of place and name entities denoting a single 
entity such as \cci{Khalifa tower}. 
\cci{Relation($e_1$,$e_2$,$r$)} creates the 
edge labeled \cci{next to} between \cci{intersection 1} and 
\cci{Khalifa tower} nodes in match 1 of Figure~\ref{fig:motiv}(b), and the 
edge labeled \cci{near} between \cci{Dubai Mall} and \cci{the building} nodes in 
match 2 of Figure~\ref{fig:motiv}(b).

\cci{Relation($r$,$e_1$,$o_1$)} creates the edge labeled \cci{prep} 
between \cci{Dubai Mall} and \cci{near} nodes in match 2 of Figure~\ref{fig:motiv}(b). 
\cci{Relation($r$,$e_2$,$o_2$)} creates the edge labeled \cci{from} 
between \cci{intersection 1} and \cci{next to} nodes in 
match 1 of Figure~\ref{fig:motiv}(b), and the 
edge labeled \cci{from this} between \cci{near} and \cci{the building} 
nodes in match 2 of Figure~\ref{fig:motiv}(b).

After constructing the relations, 
the user is interested to relate the discovered entities and relational entities
that express the same concept; in particular, the same place. 
\framework provides the \cci{isA} predicate as a default cross-reference
relation which creates the edge between the nodes \cci{Khalifa Tower} and 
\cci{The building} in  Figure~\ref{fig:intromotiv}.

>>>>

\setcode{utf8}
\setarab
\transfalse
\begin{figure*}[tb!]
\renewcommand{\arraystretch}{1.5}% Wider
\resizebox{2\columnwidth}{!}{
\begin{tabular}{cc}
\resizebox{\columnwidth}{!}{
	\begin{tabular}{c|c|c|c}
		MBF & description & formula & matches\\ \hline
        N & name of person & $category~=~Name\_of\_Person$ & $n_1,n_2,n_3$ \\ \hline
        P & name of place & $category~=~Name\_of\_Place$ & $p_1,p_2,...,p_7$ \\ \hline
        R & relative position & $stem\in$ \{\RL{قرب},\RL{في},\dots\} & $r_1,r_2,r_3,r_4$ \\ \hline
        U & numerical term & $stem\in$ \{\RL{أول},\RL{ثاني},\dots\} & $u_1,u_2$ \\
    \end{tabular}
} 
&
\resizebox{\columnwidth}{!}{
\begin{tabular}{cc}
\relsize{-2} \input{figures/match1.tex} &
\relsize{-2} \input{figures/match2.tex} \\
    \end{tabular}
} \\
\end{tabular}
}
  \caption{\framework Boolean formulae, and matches of regular expression $(P|N)+~O?~R~O\wedge 2~(P|N|U)+$}
  \label{fig:motiv}
\end{figure*}
\transtrue
\setcode{standard}

We motivate \framework with the example of Figure~\ref{fig:intromotiv} 
that shows how to extract direction entities from sample Arabic text. 
The text includes directions to a shopping mall and the 
figure presents a transliteration and an English translation of the text.
%The text contains some details that are not interesting to the direction extraction task. 
Interesting entities, such as names, places, relative directions, 
and numerical terms are highlighted with boxes in the text. 
We are interested in detecting those entities and presenting 
the direction as relational entities as 
shown in Figure~\ref{fig:intromotiv} using \framework.

@@

{\bf Entity detection.~}
As described in the table of Figure~\ref{fig:motiv}, the user denotes 
Using \framework, the user denotes the ``name of person'' entities with formula $N$ 
which requires the category feature in the morphological solution of a word
to be {\tt name of person}. 
The entities $n_1,n_2,$ and $n_3$ are matches of the formula $N$. 
Similarly, the user specifies formula $P$ to denote ``name of place'' entities. 
The user specifies formula 
$R$ to denote ``relative position'' entities, and requires the stem feature to belong 
to a selected list of 
stems containing \RL{fy} and \RL{qrb}. 
Similarly, $U$ denotes numerical terms and is a disjunction of constraints 
requiring the stem feature to belong to a set of stems such as
\RL{'wl}(first), \RL{_tAny}(second), \dots, \RL{tAs`}(ninth), and \RL{`A^sr}(tenth).

\framework calls an inhouse open source morphological 
analyzer based on finite state transducers~\cite{ZaMaColing2012DemosSarf} and computes matches of 
the formulae $N$, $P$, $R$, and $U$. 
We refer to all {\em other} words in the text that do not match 
a user defined formula as {\em null} words and we denote them by $O$. 
The resulting matches are illustrated with boxes and superscripts 
in Figure~\ref{fig:intromotiv} and are also listed in the table of 
Figure~\ref{fig:motiv}.

@@

The user tries several sequences of the above entities in the 
editor and checks their matches in the visualizer. 
Finally, the user is satisfied with the matches of an expression such as 
$(P|N)+~O?~R~O\wedge 2~(P|N|U)+$ where 
$|,+,?,$ and $\wedge k$ denote disjunction, one or more, zero or one, and
up to $k$ matches, respectively.
The expression specifies a sequence of places or names of persons, 
optionally followed by a null word, 
followed by one relative position, followed by up to two possible null words, 
followed by one or more match of name of place, name of person, or numerical term. 
$O$ and $\wedge 2$ are used in the expression to allow for flexible matches. 
The user reaches the satisfying matches by experimenting with the visualizer and 
the expression editor which do not
require knowledge and expertise in regular expressions.

\framework computes the matches of the expression in the text. 
The match trees in Figure~\ref{fig:motiv} illustrate two of them. 
The first match tree refers to the text 
\RL{brj _hlyfT bAlqrb mn AltqA.t` Al-'wl}(Khalifa Tower next to 
the first intersection). 
The second match tree refers to the text 
\RL{dby mwl `l_A mqrbT mn h_dA Almbn_A}(Dubai Mall is located near this building).
The nodes of the trees are entities and the edges and internal nodes are text, 
morphology-based, 
and word distance based relational entities. 
%The sequence and structure gives us the text of a parent node from the children nodes. 
%The same sequence can also give us the interesting (matching) morphological features. 
%The word distance is defined and abstracted by the matches of the regular expression 
%operators (internal nodes).

{\bf User defined direction relations.~} 
The user now uses the relation editor to declare user defined relations that relate 
parts of the matches of the expression with each other. 
%The aim of the semantic relations to be defined is to construct the entity-relation 
%graph shown in Figure~\ref{fig:intromotiv}. 
Intuitively, 
the user wants to create relations between places, names, and numerical entities. 
A relation between two entities can be a prepositional entity.
For example, the entities \RL{dby mwl}(Dubai Mall) and \RL{Almbn_A}(the building) are 
related by the preposition \RL{mqrbT}(near) as shown in Figure~\ref{fig:intromotiv}.

Let $e_1,o_1,r,o_2,$ and $e_2$ be the 
$(P|N)+$,$O?$,$R$,$O\wedge 2$, and $(P|N|U)+$ subexpressions, respectively.
The user selected $(P|N)+$ to be an entity after noticing in the visualizer  
that it happens to capture 
non-separated sequences of place and name entities denoting a single 
entity such as \cci{Khalifa tower}.

The declaration of \cci{Relation($e_1$,$e_2$,$r$)} creates the 
edge labeled \cci{next to} between \cci{intersection 1} and 
\cci{Khalifa tower} nodes in match 1 of Figure~\ref{fig:motiv}(b), and the 
edge labeled \cci{near} between \cci{Dubai Mall} and \cci{the building} nodes in 
match 2 of Figure~\ref{fig:motiv}(b). 
%Those relations will be constructed if $e_1$, $e_2$, and $r$ exist as match entities.
%
\cci{Relation($r$,$e_1$,$o_1$)} creates the edge labeled \cci{prep} 
between \cci{Dubai Mall} and \cci{near} nodes in match 2 of Figure~\ref{fig:motiv}(b). 
%This relation will be constructed if $r$, $e_1$, and $o_1$ exist as match entities.
%
\cci{Relation($r$,$e_2$,$o_2$)} creates the edge labeled \cci{from} 
between \cci{intersection 1} and \cci{next to} nodes in 
match 1 of Figure~\ref{fig:motiv}(b), and the 
edge labeled \cci{from this} between \cci{near} and \cci{the building} 
nodes in match 2 of Figure~\ref{fig:motiv}(b). 
%This relation will be constructed if $r$, $e_2$, and $o_2$ exist as match entities.

%\subsection*{Cross-reference}

After constructing the user defined relations, 
the user is interested to relate the discovered entities and relational entities
that express the same concept; in particular, the same place. 
\framework provides the \cci{isA} predicate as a default cross-reference
relation which creates the edge between the nodes \cci{Khalifa Tower} and 
\cci{The building} in  Figure~\ref{fig:intromotiv}.
%The \cci{isA} predicate relates an entity $w_21$ to an entity $w_2$ two entities if they are direct synonyms 
%synonymous or one entity is a synonym of a synonym of the other entity.

### 2018-03-14 tex/merf/OldSubmission/relatedwork.tex
old submission files

%SystemT\cite{chiticariu2010systemt}, 
%TEXTMARKER\cite{atzmueller2008rule}, 
%Urbain\cite{urbain2012user}, 
%QARAB\cite{hammo2002qarab}

\begin{table}[tb!]
\caption{Comaprison of \framework with SystemT, 
TEXTMARKER, 
Urbain, 
QARAB}
\resizebox{\columnwidth}{!}{
\begin{tabular}{l|c|c|c|c|c}
%{p{2.7cm}|p{.7cm}|p{1.2cm}|p{1.2cm}|p{1.2cm}|p{1.2cm}}
Features & \framework & SystemT & TEXT\linebreak MARKER & Urbain & QARAB\\ \hline
Query type & MRE & AQL & matching \linebreak rules & natural language & natural language \\ \hline
Morphology support& $\checked$ & - & - & OpenNLP & Parser \\ \hline
Relations & $\checked$ & - & - & $\checked$ & - \\ \hline
Actions & $\checked$ & - & - & - & - \\ \hline
Editor & $\checked$ & - & $\checked$ & - & - \\ \hline
Tag visualization & $\checked$ & - & $\checked$ & - & - \\ \hline
Graph visualization & $\checked$ & - & - & - & - \\
\end{tabular}
}
\label{tab:iecomp}
\end{table}

{\bf Information Extraction.~}
The common pattern specification language (CPSL) targets system independent 
IE specifications. 
CPSL consists of 
(1) a declaration part specifying names and labels, 
(2) a rule definition part specifying patterns, regular expressions and associated actions, and 
(3) and macro text substitution part~\cite{appelt1998common}.
\framework extends CPSL with Arabic morphological features, code actions, and user defined relations.

@@

DUALIST is an annotation system for quickly building classifiers for 
text processing tasks using active learning and semi-supervised learning~\cite{settles2011closing}. 
The classifier is interactive as it queries the annotator 
for entity detetction correction and annotation correction. 
\framework doesn't support classification tasks. 
However, \framework provides an interactive GUI where the user can edit MBF and MRE tags. 
This interactive environment contributes to the regular expression 
extraction and semantic relation construction which increases the overall accuracy.

WordNet is a lexical reference system that mimics human lexical memory 
%inspired by psycholinguistic theories 
and that relates words based on their semantic values and their functional 
categories: nouns, verbs, adjectives, adverbs, and function words~\cite{intro}. 
The $Syn^k$ feature in \framework is inspired by WordNet.

{\bf Annotation tools.~}
An overview of annotation tools and their Arabic-English word alignment issues 
concludes with a set of rules and guidelines needed in an Arabic annotation alignment tool
~\cite{kholidy2010towards}. 
The work in \cite{dukes2011supervised} presents a collaborative effort towards morphological 
and syntactic annotation of the Quran. 
\cite{dorr2010interlingual} presents a framework for interlingual 
annotation of parallel text corpora with multi-level representations. 
\cite{kulick2010consistent} presents the integration of the 
Standard Arabic Morphological Analyzer (SAMA) into the workflow of the Arabic Treebank.

MMAX2 is a manual multi-level linguistic annotation tool with an XML 
based data model~\cite{mmax2}. 
It enables the user to create, browse, visualize, and query annotations
and may be able to resolve coreference tags. 
BRAT~\cite{brat} and WordFreak~\cite{wordfreak} are manual 
multi-lingual user friendly web-based annotators that allow the construction 
of entity and relation annotation corpora~\cite{brat}. 
They can be extended through plug-ins to enable automatic annotators 
and customized annotation visualization specifications.
Knowtator~\cite{ogren2006knowtator} is a general purpose incremental text annotation tool 
implemented as a Prot\'eg\'e~\cite{gennari2003evolution} plug-in. 
Prot\'eg\'e is an open-source platform with a suite of tools to construct domain 
models and knowledge-based applications with ontologies. 
However, it doesn't support the Arabic language.

\framework differs from MMAX2, BRAT, WordFreak, and Knowtator in that it is 
an automatic annotator that allows manual corrections 
and sophisticated tag type and relation specifications over
Arabic morphological features.

@@

Task specific annotation tools such as~\cite{alrahabi2006semantic}
uses enunciation semantic maps to automatically annotate 
directly reported Arabic and French speech. 
AraTation is another task specific tool for semantic annotation of 
Arabic news using web ontology based semantic maps~\cite{saleh2009aratation}.
We differ in that \framework is general, and not task specific, and it uses 
morphology-based features as atomic terms.
Fassieh is a commercial Arabic text annotation tool that enables the production of large 
Arabic text corpora~\cite{attia2009fassieh}. 
The tool supports Arabic text factorization including morphological analysis, POS tagging, 
full phonetic transcription, and lexical semantics analysis in an automatic mode. 
Fassieh is not directly accessible to the research community and requires commercial licensing. 
\framework is open source and differs in that it allows the user to build tag types 
and extract entities and relations from text.

>>>>

\begin{table}[tb!]
\caption{Comaprison of IE system}
\resizebox{\columnwidth}{!}{
\begin{tabular}{p{1.4cm}|p{.7cm}|p{1.2cm}|p{1.2cm}|p{1.2cm}|p{1.2cm}}
Features & \framework & SystemT\cite{chiticariu2010systemt} & TEXT\cite{atzmueller2008rule} \linebreak MARKER
& Urbain\cite{urbain2012user} & QARAB\cite{hammo2002qarab}\\ \hline
Query type & MRE & AQL & matching \linebreak rules & natural language & natural language \\ \hline
Morphology support& $\checked$ & - & - & OpenNLP & Parser \\ \hline
Relations & $\checked$ & - & - & $\checked$ & - \\ \hline
Actions & $\checked$ & - & - & - & - \\ \hline
Editor & $\checked$ & - & $\checked$ & - & - \\ \hline
Tag visualization & $\checked$ & - & $\checked$ & - & - \\ \hline
Graph visualization & $\checked$ & - & - & - & - \\
\end{tabular}
}
\label{tab:iecomp}
\end{table}

{\bf Information Extraction.~}
The common pattern specification grammar language (CPSL) targets system independent 
IE specifications. 
CPSL consists of 
(1) a declaration part specifying names and labels, 
(2) a rule definition part specifying patterns, regular expressions and associated actions, and 
(3) and macro text substitution part~\cite{appelt1998common}.
\framework extends CPSL with Arabic morphological features, code actions, and user defined relations.

@@

DUALIST is an annotation system for quickly building classifiers for 
text processing tasks using active learning and semi-supervised learning~\cite{settles2011closing}. 
The classifier is interactive as it queries the annotator 
for entity detetction correction and annotation correction. 
\framework doesn't support classification tasks. 
However, \framework provides an interactive GUI where the user can edit MBF and MRE tags. 
This interactive environment contributes to the regular expression 
extraction and semantic relation construction which increases the overall accuracy of the framework.

WordNet is a lexical reference system that mimics human lexical memory inspired 
by psycholinguistic theories and that relates words based on their semantic values and their functional 
categories: nouns, verbs, adjectives, adverbs, and function words~\cite{intro}. 
%The lexicon is composed of five categories: nouns, verbs, adjectives, adverbs, and function words; 
%each category with its distinct organization to better represent the psychological complexity of lexical knowledge~\cite{nouns,adjectives}. 
The $Syn^k$ feature in \framework is inspired by WordNet.

{\bf Annotation tools.~}
An overview of annotation tools and their Arabic-English word alignment issues 
concludes with a set of rules and guidelines needed in an Arabic annotation alignment tool
~\cite{kholidy2010towards}. 
The work in \cite{dukes2011supervised} presents a collaborative effort towards morphological 
and syntactic annotation of the Quran. 
%The task is held through online supervised collaboration using a multi-stage approach that 
%includes automatic POS tagging, manual verification, and online supervised collaborative 
%proofreading. 
The work in \cite{dorr2010interlingual} presents a framework for interlingual 
annotation of parallel text corpora with multi-level representations. 
%This corpora is important for NLP tasks including text summarization, information retrieval, 
%and machine translation. 
The work in \cite{kulick2010consistent} presents the integration of the 
Standard Arabic Morphological Analyzer (SAMA) into the workflow of the Arabic Treebank.
%Such tasks motivated us to build \framework, a morphology-based open source annotation tool 
%for the Arabic language.

MMAX2 is a manual multi-level linguistic annotation tool with an XML 
based data model~\cite{mmax2}. 
It enables the user to create, browse, visualize, and query annotations
and may be able to resolve coreference tags. 
BRAT~\cite{brat} and WordFreak~\cite{wordfreak} are manual 
multi-lingual user friendly web-based annotators that allow the construction 
of entity and relation annotation corpora~\cite{brat}. 
%BRAT provides an API for automated annotators to provide annotations. 
They can be extended through plug-ins to enable automatic annotators 
and customized annotation visualization specifications.
%AGTK is a toolkit for the development of text and speech annotation tools~\cite{agtk}. 
%It provides import APIs from other data and graphical user interface (GUI) components. 
Knowtator~\cite{ogren2006knowtator} is a general purpose incremental text annotation tool 
implemented as a Prot\'eg\'e~\cite{gennari2003evolution} plug-in. 
Prot\'eg\'e is a free, open-source platform with a suite of tools to construct domain 
models and knowledge-based applications with ontologies~\cite{gennari2003evolution}. 
However, Prot\'eg\'e doesn't support the Arabic language.

\framework differs from MMAX2, BRAT, WordFreak, and Knowtator in that it is 
an automatic annotator that allows manual corrections 
and sophisticated tag type and relation specifications using regular expressions over
Arabic morphological features.

@@

Task specific annotation tools such as~\cite{alrahabi2006semantic}
uses enunciation semantic maps to automatically annotate 
directly reported Arabic and French speech. 
AraTation is another task specific tool for semantic annotation of 
Arabic news using web ontology based semantic maps~\cite{saleh2009aratation}.
We differ in that \framework is general, and not task specific, and it uses 
morphology-based features as atomic terms.
Fassieh is a commercial Arabic text annotation tool that enables the production of large 
Arabic text corpora~\cite{attia2009fassieh}. 
The tool supports Arabic text factorization including morphological analysis, POS tagging, 
full phonetic transcription, and lexical semantics analysis in an automatic mode. 
Fassieh is not directly accessible to the research community and requires commercial licensing. 
\framework is open source and differs in that it allows the user to build tag types 
and extract entities and relations from text.
%using Boolean formulae of several atomic terms and define regular expressions based on the Boolean formulae.
